---
title: "Rates_std"
author: "Fabian Sanabria-Mora"
format: html
editor: source
---

# Direct rate standardisation (by age)

This Quarto document presents the process of direct standardisation of suicide rates using the age-based method and the **WHO standard population** (see [WHO_std](https://cdn.who.int/media/docs/default-source/gho-documents/global-health-estimates/gpe_discussion_paper_series_paper31_2001_age_standardization_rates.pdf)). The first step was to upload the database of suicides (2004 - 2022).

```{r Upload}

setwd("")
sui <- suppressWarnings(readxl::read_xlsx("datasui_0422.xlsx"))

#Rename variables
library(dplyr)
sui <- sui %>% rename(year = año, age = edad, month = mes_sui, age_group = grupo_edad, sex = sexo, sui_zone = zona_sui)

```

## Step 1: Columns

Columns not used for standardisation and further analysis are removed. Subcategories are renamed and variables are revised.

```{r Columns}

sui <- sui %>% dplyr::select(-estado_civil, -nacionalidad, -escolaridad, -ocupacion, -dia_sui, -rango_3h, -rango_6h, -municipio, -departamento, -escenario, -razon_sui, -ancestro_raza, -mecanismo)

#Rename
sui <- sui %>%
  mutate(
    sex = recode(sex, 
                 "Mujer" = "Women",
                 "Hombre" = "Men",
                 "Sin información" = "No information"),
    
    month = recode(month,
                   "Enero" = "January",
                   "Febrero" = "February",
                   "Marzo" = "March",
                   "Abril" = "April",
                   "Mayo" = "May",
                   "Junio" = "June",
                   "Julio" = "July",
                   "Agosto" = "August",
                   "Septiembre" = "September",
                   "Octubre" = "October",
                   "Noviembre" = "November",
                   "Diciembre" = "December",
                   "Sin información" = "No information"),
    
    sui_zone = recode(sui_zone,
                      "Cabecera municipal" = "Urban",
                      "Centro poblado (corregimiento, inspección de policía y caserío)" = "Rural", 
                      "Parte rural (vereda y campo)" = "Rural",
                      "Sin información" = "No information"),
    
    age_group = recode(age_group,
                       "Sin información" = "No information"),
  
  )



#Class "factor" and ordered
sui <- sui %>%
  mutate(
    sex = factor(sex,
                 levels = c("Women", "Men", "No information"), ordered = T),
    
    month = factor(month,
                   levels = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December", "No information"), ordered = T),
    
    sui_zone = factor(sui_zone,
                      levels = c("Rural", "Urban", "No information"), ordered = T),
    
  )

unique(sui$year)
class(sui$year)

unique(sui$month)
class(sui$month)

unique(sui$age)
class(sui$age)

unique(sui$age_group)
class(sui$age_group)

unique(sui$sex)
class(sui$sex)

unique(sui$sui_zone)
class(sui$sui_zone)

```

Note: *"Centro poblado"* and *"Parte rural"* are now called *"Rural"*.

## Step 2: Age

Ages are grouped according to standardisation categories.

```{r Age}

#Delete < 5 years old
sui1 <- sui %>% filter(age != "0" | age != "1" | age != "2" | age != "3" | age != "4")
unique(sui1$age)

#5 - 75 and more (by decades) 
library(epikit)
sui_dec <- sui1 %>%
  mutate(age_cat = age_categories(age, breakers = c(5, 15, 25, 35, 45, 55, 65, 75)))
table(sui_dec$age_cat, useNA = "always")

#as factor
sui_dec <- sui_dec %>%
  mutate(age_cat = factor(age_cat, 
    levels = c("5-14", "15-24", "25-34", "35-44", "45-54", "55-64", "65-74", "75+"), ordered = T))

#Delete variables
sui_dec <- sui_dec %>% dplyr::select(-age, -age_group, -id)



#5 - 60 and more (by DANE cycles)
sui_cyc <- sui1 %>%
  mutate(age_cat = age_categories(age, breakers = c(5, 18, 45, 60)))
table(sui_cyc$age_cat, useNA = "always")

#as factor
sui_cyc <- sui_cyc %>%
  mutate(age_cat = factor(age_cat, 
    levels = c("5-17", "18-44", "45-59", "60+"), ordered = T))

#Delete variables
sui_cyc <- sui_cyc %>% dplyr::select(-age, -age_group, -id)

```

## Step 3: Cases

Suicide cases are calculated stratified by the other variables.

```{r Cases}

#by decades
sui_dec <- sui_dec %>%
  group_by(year, month, sex, sui_zone, age_cat) %>%
  summarise(cases = n(), .groups = "drop")

#by cycles
sui_cyc <- sui_cyc %>%
  group_by(year, month, sex, sui_zone, age_cat) %>%
  summarise(cases = n(), .groups = "drop")


#distribution of "cases"
library(ggplot2)
cases_dist <- ggplot(sui_cyc, aes(x=cases)) + theme_bw(16) + 
            geom_histogram(aes(y=..density..),      
                           binwidth=1,
                           colour="black", 
                           fill="lightgreen") +
            geom_density(alpha=.2, fill="#FF6666")  
cases_dist

#Gamma parameters
MASS::fitdistr(sui_cyc$cases, "Gamma")

#graph
cases_dist + stat_function(data=sui_cyc, 
                        fun = dgamma, 
                        args=list(shape=0.856898696, 
                                  rate=0.077203935),
                        colour="darkred", 
                        size=1.1) + 
  stat_function(data=sui_cyc, 
                fun = dlnorm, 
                args=list(meanlog=mean(log(sui_cyc$cases+1)),
                          sdlog=sd(log(sui_cyc$cases+1))),
                colour="blue", 
                lty=5,
                size=1.1)  +  
  xlim(c(-1, 60)) +
  labs(title = "Gamma + log-normal distribution")

#Conclusion: Since the interest is to capture the small amount of data observed on the left better, the gamma distribution fits better. This must be taken into account when calculating confidence intervals for suicide incidence rates.

#The code was modified from: https://github.com/devanmcg/IntroRangeR/tree/master/07_DataDistributions

```

Note that `summarise(cases = n(), .groups = "drop")` collapses the database by counting the repetitions of each combination.

## Step 4: Prepare databases (DANE and WHO)

### By decades

```{r Prepare databases / by decades}

#DANE
dane <- suppressWarnings(readxl::read_xlsx("proyecnacional_2003_2023.xlsx"))

#Rename and recode variables
library(dplyr)
dane_dec <- dane %>% rename(year = "AÑO", sui_zone = "ÁREA GEOGRÁFICA")
dane_dec <- dane_dec %>%
  mutate(sui_zone = recode(sui_zone, "Cabecera" = "Urban", "Centros Poblados y Rural Disperso" = "Rural"))

#Delete age < 5 and "Total"
dane_dec <- dane_dec %>%
  dplyr::select(-Hombres_0, -Hombres_1, -Hombres_2, -Hombres_3, -Hombres_4, -Mujeres_0, -Mujeres_1, -Mujeres_2, -Mujeres_3, -Mujeres_4)

dane_dec <- dane_dec %>%
  dplyr::select(-(starts_with("Total")))

#Sum by categories
#Men
dane_dec <- dane_dec %>%
  rowwise() %>%
  mutate("Men_5-14" = sum(c_across(Hombres_5:Hombres_14))) %>%
  ungroup()

dane_dec <- dane_dec %>%
  rowwise() %>%
  mutate("Men_15-24" = sum(c_across(Hombres_15:Hombres_24))) %>%
  ungroup()

dane_dec <- dane_dec %>%
  rowwise() %>%
  mutate("Men_25-34" = sum(c_across(Hombres_25:Hombres_34))) %>%
  ungroup()

dane_dec <- dane_dec %>%
  rowwise() %>%
  mutate("Men_35-44" = sum(c_across(Hombres_35:Hombres_44))) %>%
  ungroup()

dane_dec <- dane_dec %>%
  rowwise() %>%
  mutate("Men_45-54" = sum(c_across(Hombres_45:Hombres_54))) %>%
  ungroup()

dane_dec <- dane_dec %>%
  rowwise() %>%
  mutate("Men_55-64" = sum(c_across(Hombres_55:Hombres_64))) %>%
  ungroup()

dane_dec <- dane_dec %>%
  rowwise() %>%
  mutate("Men_65-74" = sum(c_across(Hombres_65:Hombres_74))) %>%
  ungroup()

dane_dec <- dane_dec %>%
  rowwise() %>%
  mutate("Men_75+" = sum(c_across(Hombres_75:Hombres_100))) %>%
  ungroup()

dane_dec <- dane_dec %>%
  dplyr::select(-(starts_with("Hombres")))


#Women
dane_dec <- dane_dec %>%
  rowwise() %>%
  mutate("Women_5-14" = sum(c_across(Mujeres_5:Mujeres_14))) %>%
  ungroup()

dane_dec <- dane_dec %>%
  rowwise() %>%
  mutate("Women_15-24" = sum(c_across(Mujeres_15:Mujeres_24))) %>%
  ungroup()

dane_dec <- dane_dec %>%
  rowwise() %>%
  mutate("Women_25-34" = sum(c_across(Mujeres_25:Mujeres_34))) %>%
  ungroup()

dane_dec <- dane_dec %>%
  rowwise() %>%
  mutate("Women_35-44" = sum(c_across(Mujeres_35:Mujeres_44))) %>%
  ungroup()

dane_dec <- dane_dec %>%
  rowwise() %>%
  mutate("Women_45-54" = sum(c_across(Mujeres_45:Mujeres_54))) %>%
  ungroup()

dane_dec <- dane_dec %>%
  rowwise() %>%
  mutate("Women_55-64" = sum(c_across(Mujeres_55:Mujeres_64))) %>%
  ungroup()

dane_dec <- dane_dec %>%
  rowwise() %>%
  mutate("Women_65-74" = sum(c_across(Mujeres_65:Mujeres_74))) %>%
  ungroup()

dane_dec <- dane_dec %>%
  rowwise() %>%
  mutate("Women_75+" = sum(c_across(Mujeres_75:Mujeres_100))) %>%
  ungroup()

dane_dec <- dane_dec %>%
  dplyr::select(-(starts_with("Mujeres")))

#Pivot data (wide to long)
library(tidyr)
dane_dec <- dane_dec %>%
  pivot_longer(cols = `Men_5-14`:`Women_75+`, names_to = c("sex", "age_cat"), names_sep = "_", values_to = "pop")

#Calculate the differences between the populations of all years
dane_dec_diff <- dane_dec %>%
  arrange(sex, age_cat, sui_zone, year) %>% #order by groups
  group_by(sex, age_cat, sui_zone) %>% 
  mutate(diff = pop - lag(pop)) %>% #subtract the value from the previous year
  ungroup()

#Divide by 12
dane_dec_diff <- dane_dec_diff %>%
  mutate(pop_m = diff/12)

#Expand by months
dane_dec_months <- dane_dec_diff %>%
  mutate(month = NA) %>% #create "month" variable
  complete(nesting(year, sex, age_cat, sui_zone, pop, pop_m), month = 1:12) #For each combination of variables, assign the rows of the 12 months

#Now, in the variable "pop1" the differences in the populations will be added month by month ("pop_m"), from January to December
dane_dec_months <- dane_dec_months %>%
  mutate(pop1 = 0)

dane_dec_months <- dane_dec_months %>%
  arrange(year, sex, age_cat, sui_zone, month) %>%
  group_by(year, sex, age_cat, sui_zone) %>%
  mutate(
    pop1 = pop + cumsum(pop_m) - pop_m
  ) %>% #subtracting "pop_m" prevents January from being added twice
  ungroup()

#Some adjustments
dane_dec_months <- dane_dec_months %>%
  dplyr::select(-pop, -pop_m, -diff)

dane_dec_months <- dane_dec_months %>%
  filter(year != "2003" & year != "2023")

dane_dec_months <- dane_dec_months %>%
  filter(month != "")

dane_dec_months <- dane_dec_months %>%
  mutate(
    month = recode(month, 
                 "1" = "January",
                 "2" = "February",
                 "3" = "March",
                 "4" = "April",
                 "5" = "May",
                 "6" = "June",
                 "7" = "July",
                 "8" = "August",
                 "9" = "September",
                 "10" = "October",
                 "11" = "November",
                 "12" = "December")
    )


#WHO (import from STATA.dta)
world_pop_decades <- haven::read_dta("who_pop_5to75+.dta")

world_pop_decades <- world_pop_decades %>%
  mutate(grupo_edad = factor(grupo_edad, 
    levels = c("1", "2", "3", "4", "5", "6", "7", "8"), ordered = T)) #convert to factor

world_pop_decades <- world_pop_decades %>%
  mutate(
    grupo_edad = recode(grupo_edad, 
                 `1` = "5-14",
                 `2` = "15-24",
                 `3` = "25-34",
                 `4` = "35-44",
                 `5` = "45-54",
                 `6` = "55-64",
                 `7` = "65-74",
                 `8` = "75+")
  )

world_pop_decades <- world_pop_decades %>%
  rename(age_cat = "grupo_edad") 

world_pop_decades <- world_pop_decades %>%
  rename(pop = "population") 

#as dataframe
world_pop_decades <- as.data.frame(world_pop_decades)


```

-   DANE's population projections for 2020, 2021 and 2022 are adjusted for the impact of the COVID-19 pandemic.
-   Note that `rowwise()` allows operations per row.
-   `names_to` creates the new variables.
-   `names_sep` is the term that separates the two variables.
-   `values_to` creates the column of values ("pop").

### By cycles

```{r Prepare databases / by cycles}

#DANE
dane <- suppressWarnings(readxl::read_xlsx("proyecnacional_2003_2023.xlsx"))

#Rename and recode variables
library(dplyr)
dane_cyc <- dane %>% rename(year = "AÑO", sui_zone = "ÁREA GEOGRÁFICA")
dane_cyc <- dane_cyc %>%
  mutate(sui_zone = recode(sui_zone, "Cabecera" = "Urban", "Centros Poblados y Rural Disperso" = "Rural"))

#Delete age < 5 and "Total"
dane_cyc <- dane_cyc %>%
  dplyr::select(-Hombres_0, -Hombres_1, -Hombres_2, -Hombres_3, -Hombres_4, -Mujeres_0, -Mujeres_1, -Mujeres_2, -Mujeres_3, -Mujeres_4)

dane_cyc <- dane_cyc %>%
  dplyr::select(-(starts_with("Total")))

#Sum by categories
#Men
dane_cyc <- dane_cyc %>%
  rowwise() %>%
  mutate("Men_5-17" = sum(c_across(Hombres_5:Hombres_17))) %>%
  ungroup()

dane_cyc <- dane_cyc %>%
  rowwise() %>%
  mutate("Men_18-44" = sum(c_across(Hombres_18:Hombres_44))) %>%
  ungroup()

dane_cyc <- dane_cyc %>%
  rowwise() %>%
  mutate("Men_45-59" = sum(c_across(Hombres_45:Hombres_59))) %>%
  ungroup()

dane_cyc <- dane_cyc %>%
  rowwise() %>%
  mutate("Men_60+" = sum(c_across(Hombres_60:Hombres_100))) %>%
  ungroup()

dane_cyc <- dane_cyc %>%
  dplyr::select(-(starts_with("Hombres")))


#Women
dane_cyc <- dane_cyc %>%
  rowwise() %>%
  mutate("Women_5-17" = sum(c_across(Mujeres_5:Mujeres_17))) %>%
  ungroup()

dane_cyc <- dane_cyc %>%
  rowwise() %>%
  mutate("Women_18-44" = sum(c_across(Mujeres_18:Mujeres_44))) %>%
  ungroup()

dane_cyc <- dane_cyc %>%
  rowwise() %>%
  mutate("Women_45-59" = sum(c_across(Mujeres_45:Mujeres_59))) %>%
  ungroup()

dane_cyc <- dane_cyc %>%
  rowwise() %>%
  mutate("Women_60+" = sum(c_across(Mujeres_60:Mujeres_100))) %>%
  ungroup()

dane_cyc <- dane_cyc %>%
  dplyr::select(-(starts_with("Mujeres")))

#Pivot data (wide to long)
library(tidyr)
dane_cyc <- dane_cyc %>%
  pivot_longer(cols = `Men_5-17`:`Women_60+`, names_to = c("sex", "age_cat"), names_sep = "_", values_to = "pop")

#Calculate the differences between the populations of all years
dane_cyc_diff <- dane_cyc %>%
  arrange(sex, age_cat, sui_zone, year) %>% #order by groups
  group_by(sex, age_cat, sui_zone) %>% 
  mutate(diff = pop - lag(pop)) %>% #subtract the value from the previous year
  ungroup()

#Divide by 12
dane_cyc_diff <- dane_cyc_diff %>%
  mutate(pop_m = diff/12)

#Expand by months
dane_cyc_months <- dane_cyc_diff %>%
  mutate(month = NA) %>% #create "month" variable
  complete(nesting(year, sex, age_cat, sui_zone, pop, pop_m), month = 1:12) #For each combination of variables, assign the rows of the 12 months

#Now, in the variable "pop1" the differences in the populations will be added month by month ("pop_m"), from January to December
dane_cyc_months <- dane_cyc_months %>%
  mutate(pop1 = 0)

dane_cyc_months <- dane_cyc_months %>%
  arrange(year, sex, age_cat, sui_zone, month) %>%
  group_by(year, sex, age_cat, sui_zone) %>%
  mutate(
    pop1 = pop + cumsum(pop_m) - pop_m
  ) %>% #subtracting "pop_m" prevents January from being added twice
  ungroup()

#Some adjustments
dane_cyc_months <- dane_cyc_months %>%
  dplyr::select(-pop, -pop_m, -diff)

dane_cyc_months <- dane_cyc_months %>%
  filter(year != "2003" & year != "2023")

dane_cyc_months <- dane_cyc_months %>%
  filter(month != "")

dane_cyc_months <- dane_cyc_months %>%
  mutate(
    month = recode(month, 
                 "1" = "January",
                 "2" = "February",
                 "3" = "March",
                 "4" = "April",
                 "5" = "May",
                 "6" = "June",
                 "7" = "July",
                 "8" = "August",
                 "9" = "September",
                 "10" = "October",
                 "11" = "November",
                 "12" = "December")
    )


#WHO (import from STATA.dta)
world_pop_cyc <- haven::read_dta("who_pop_cycle.dta")

world_pop_cyc <- world_pop_cyc %>%
  mutate(grupo_edad = factor(grupo_edad, 
    levels = c("1", "2", "3", "4"), ordered = T)) #convert to factor

world_pop_cyc <- world_pop_cyc %>%
  mutate(
    grupo_edad = recode(grupo_edad, 
                 `1` = "5-17",
                 `2` = "18-44",
                 `3` = "45-59",
                 `4` = "60+")
  )

world_pop_cyc <- world_pop_cyc %>%
  rename(age_cat = "grupo_edad") 

world_pop_cyc <- world_pop_cyc %>%
  rename(pop = "population") 

#as dataframe
world_pop_cyc <- as.data.frame(world_pop_cyc)


```

## Step 5: Merge

This step consists of merging this database with population projections (DANE) and the standard population (WHO).

```{r Merge}

#by decades
dec_def <- dplyr::left_join(sui_dec, dane_dec_months, by = c("year", "month", "sex", "sui_zone", "age_cat"))

dec_def1 <- dplyr::left_join(dec_def, world_pop_decades, by = "age_cat")

#If there are missing values in a vector, the "dsr" command does not work, so these must be removed
dec_def1 <- dec_def1 %>%
  filter(pop1 != "")
#5732

#as dataframe
dec_def1 <- as.data.frame(dec_def1)

#as factor (this is required for the "dsr" function)
dec_def1 <- dec_def1 %>%
  mutate(
    sex = factor(sex,
                 levels = c("Women", "Men"), ordered = T),
    
    month = factor(month, 
                   levels = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December", "No information"), ordered = T),
    
    sui_zone = factor(sui_zone, 
                      levels = c("Rural", "Urban"), ordered = T),
    
    age_cat = factor(age_cat, 
                     levels = c("5-14", "15-24", "25-34", "35-44", "45-54", "55-64", "65-74", "75+"), ordered = T),
    
    year = factor(year,
                  levels = c("2004", "2005", "2006", "2007", "2008", "2009", "2010", "2011", "2012", "2013", "2014", "2015", "2016", "2017", "2018", "2019", "2020", "2021", "2022"), ordered = T)
    
  )


#by cycles
cyc_def <- dplyr::left_join(sui_cyc, dane_cyc_months, by = c("year", "month", "sex", "sui_zone", "age_cat"))

cyc_def1 <- dplyr::left_join(cyc_def, world_pop_cyc, by = "age_cat")

#If there are missing values in a vector, the "dsr" command does not work, so these must be removed
cyc_def1 <- cyc_def1 %>%
  filter(pop1 != "")
#3284

#as dataframe
cyc_def1 <- as.data.frame(cyc_def1)

#as factor (this is required for the "dsr" function)
cyc_def1 <- cyc_def1 %>%
  mutate(
    sex = factor(sex,
                 levels = c("Women", "Men"), ordered = T),
    
    month = factor(month, 
                   levels = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December", "No information"), ordered = T),
    
    sui_zone = factor(sui_zone, 
                      levels = c("Rural", "Urban"), ordered = T),
    
    age_cat = factor(age_cat, 
                     levels = c("5-17", "18-44", "45-59", "60+"), ordered = T),
    
    year = factor(year,
                  levels = c("2004", "2005", "2006", "2007", "2008", "2009", "2010", "2011", "2012", "2013", "2014", "2015", "2016", "2017", "2018", "2019", "2020", "2021", "2022"), ordered = T)
    
  )


```

## Step 6: Standardisation

Three R packages were identified to perform the standardization: `dsrTest`, `dsr` and `popEpi`. With `dsrTest`, it was not possible to calculate the rates stratified by sex and zone. With the `dsr` package, rates and CIs can be calculated (using the gamma method), nevertheless, to perform the stratification by sex and zone, it is necessary to generate **interactions** between the variables, which *apparently* modifies the results of the rates. Because of this, it was decided to use the `popEpi` package, which allows for appropriate stratification; however *"The confidence intervals are based on the normal approximation of the logarithm of the rate.The variance of the log rate that is used to derive the confidence intervals is derived using the delta method".* Because of this, confidence intervals were calculated manually by adjusting for the gamma distribution **(see step 7)**.

It should be considered that: *"If the stratum-specific incidences of a population are standardized to that population's distribution, then the standardized incidence will be equal to the crude incidence for that population. Thus, a crude incidence is an average weighted by the distribution of the study population itself. A standardized incidence can be interpreted as the crude incidence in a population that has the same stratum-specific incidences as those of the observed population, but in which the distribution of the stratiﬁcation variables is given by the standard"* (Greenland S, Rothman KJ (2008) Introduction to Stratified Analysis. In: Modern Epidemiology. Wolters Kluwer - Lippincott Williams & Wilkins, Philadelphia (USA)).

Additionally: - **dsr package** does not allow stratification by more than one subgroup, so `interaction` variables must be created depending on the interest. - Column 1, which corresponds to the **interactions** between year, month and another variable, must be separated into three new columns (`separate` function of `tidyr` package); this is necessary to adjust the format of the final tables.

### By decades

```{r by decades}


#Example with dsr package
#dsr package (url, no CRAN)
#install.packages("frailtypack")
package_url <- "https://cran.r-project.org/src/contrib/Archive/dsr/dsr_0.2.2.tar.gz"
#install.packages(package_url, repos = NULL, type = "source")

library(dsr)
#Year
rates_year_dec <- dsr(
  data = dec_def1, 
  event = cases,
  fu = pop1,
  subgroup = year,
  age_cat,
  refdata = world_pop_decades,
  method = "gamma",
  sig = 0.95,
  mp = 100000,
  decimals = 4
)

year_dec <- knitr::kable(rates_year_dec, format = "html", caption = "Suicide rates by year (per 100000)", col.names = c('year', 'suicides', 'population', 'crude_rate', '95% LCL (crude)', '95% UCL (crude)', 'std_rate', '95% LCL (std)', '95% UCL (std)'), align = 'r')
year_dec


#Year and month (global)
dec_def1$year_month <- interaction(dec_def1$year, dec_def1$month, sep = "/")

rates_ymonth_dec <- dsr(
  data = dec_def1, 
  event = cases,
  fu = pop1,
  subgroup = year_month,
  age_cat,
  refdata = world_pop_decades,
  method = "gamma",
  sig = 0.95,
  mp = 100000,
  decimals = 4
)

##New columns
rates_ymonth_dec <- rates_ymonth_dec %>%
  separate("Subgroup", into = c("year", "month"), sep = "/")
rates_ymonth_dec



#With popEpi
library(popEpi)

##Vector for weights population (order matters / by age_cat)
pop_v <- c(17290, 16690, 15540, 13740, 11410, 8270, 5170, 3065)

##overall
rate_dec <- rate(data = dec_def1, obs = cases, pyrs = pop1, adjust = age_cat, weights = pop_v)

##*100000
rate_dec <- rate_dec %>%
  mutate(across(c(rate.adj, SE.rate.adj, rate.adj.lo, rate.adj.hi, rate, SE.rate, rate.lo, rate.hi), ~ . * 100000))
rate_dec


##year and month
rates_yearm_dec <- rate(data = dec_def1, obs = cases, pyrs = pop1, print = list(year, month), adjust = age_cat, weights = pop_v)

##delete rows if "suicides" == 0
library(dplyr)
rates_yearm_dec <- rates_yearm_dec %>%
  filter(cases != 0)
rates_yearm_dec

##*100000
rates_yearm_dec <- rates_yearm_dec %>%
  mutate(across(c(rate.adj, SE.rate.adj, rate.adj.lo, rate.adj.hi, rate, SE.rate, rate.lo, rate.hi), ~ . * 100000))
rates_yearm_dec

##table
yearm_dec <- knitr::kable(rates_yearm_dec, format = "html", caption = "Suicide rates by year and month (per 100000)", col.names = c('year', 'month', 'suicides', 'population', 'std_rate', 'SE std_rate', '95% LCL (std)', '95% UCL (std)', 'crude_rate', 'SE crude_rate', '95% LCL (crude)', '95% UCL (crude)'), align = 'r')
yearm_dec


##year, month and sex
rates_yearms_dec <- rate(data = dec_def1, obs = cases, pyrs = pop1, print = list(year, month, sex), adjust = age_cat, weights = pop_v)

##delete rows if "suicides" == 0
library(dplyr)
rates_yearms_dec <- rates_yearms_dec %>%
  filter(cases != 0)
rates_yearms_dec

##*100000
rates_yearms_dec <- rates_yearms_dec %>%
  mutate(across(c(rate.adj, SE.rate.adj, rate.adj.lo, rate.adj.hi, rate, SE.rate, rate.lo, rate.hi), ~ . * 100000))
rates_yearms_dec

##table
yearms_dec <- knitr::kable(rates_yearms_dec, format = "html", caption = "Suicide rates by year, month and sex (per 100000)", col.names = c('year', 'month', 'sex', 'suicides', 'population', 'std_rate', 'SE std_rate', '95% LCL (std)', '95% UCL (std)', 'crude_rate', 'SE crude_rate', '95% LCL (crude)', '95% UCL (crude)'), align = 'r')
yearms_dec


##year, month and zone
rates_yearmz_dec <- rate(data = dec_def1, obs = cases, pyrs = pop1, print = list(year, month, sui_zone), adjust = age_cat, weights = pop_v)

##delete rows if "suicides" == 0
library(dplyr)
rates_yearmz_dec <- rates_yearmz_dec %>%
  filter(cases != 0)
rates_yearmz_dec

##*100000
rates_yearmz_dec <- rates_yearmz_dec %>%
  mutate(across(c(rate.adj, SE.rate.adj, rate.adj.lo, rate.adj.hi, rate, SE.rate, rate.lo, rate.hi), ~ . * 100000))
rates_yearmz_dec

##table
yearmz_dec <- knitr::kable(rates_yearmz_dec, format = "html", caption = "Suicide rates by year, month and zone (per 100000)", col.names = c('year', 'month', 'zone', 'suicides', 'population', 'std_rate', 'SE std_rate', '95% LCL (std)', '95% UCL (std)', 'crude_rate', 'SE crude_rate', '95% LCL (crude)', '95% UCL (crude)'), align = 'r')
yearmz_dec


##year, month and age (crudes)
rates_yearma_dec <- rate(data = dec_def1, obs = cases, pyrs = pop1, print = list(year, month, age_cat))
rates_yearma_dec

##*100000
rates_yearma_dec <- rates_yearma_dec %>%
  mutate(across(c(rate, SE.rate, rate.lo, rate.hi), ~ . * 100000))
rates_yearma_dec

##table
yearma_dec <- knitr::kable(rates_yearma_dec, format = "html", caption = "Suicide rates by year, month and age groups (per 100000)", col.names = c('year', 'month', 'age', 'suicides', 'population', 'std_rate', 'SE std_rate', '95% LCL (std)', '95% UCL (std)'), align = 'r')
yearma_dec


```

### By cycles

```{r by cycles}

##Vector for weights population (order matters / by age_cat)
pop_vc <- c(22394, 40863, 15960, 11950)

##overall
rate_cyc <- rate(data = cyc_def1, obs = cases, pyrs = pop1, adjust = age_cat, weights = pop_vc)

##*100000
rate_cyc <- rate_cyc %>%
  mutate(across(c(rate.adj, SE.rate.adj, rate.adj.lo, rate.adj.hi, rate, SE.rate, rate.lo, rate.hi), ~ . * 100000))
rate_cyc


##year and month
rates_yearm_cyc <- rate(data = cyc_def1, obs = cases, pyrs = pop1, print = list(year, month), adjust = age_cat, weights = pop_vc)

##delete rows if "suicides" == 0
library(dplyr)
rates_yearm_cyc <- rates_yearm_cyc %>%
  filter(cases != 0)
rates_yearm_cyc

##*100000
rates_yearm_cyc <- rates_yearm_cyc %>%
  mutate(across(c(rate.adj, SE.rate.adj, rate.adj.lo, rate.adj.hi, rate, SE.rate, rate.lo, rate.hi), ~ . * 100000))
rates_yearm_cyc

##table
yearm_cyc <- knitr::kable(rates_yearm_cyc, format = "html", caption = "Suicide rates by year and month (per 100000)", col.names = c('year', 'month', 'suicides', 'population', 'std_rate', 'SE std_rate', '95% LCL (std)', '95% UCL (std)', 'crude_rate', 'SE crude_rate', '95% LCL (crude)', '95% UCL (crude)'), align = 'r')
yearm_cyc


##year, month and sex
rates_yearms_cyc <- rate(data = cyc_def1, obs = cases, pyrs = pop1, print = list(year, month, sex), adjust = age_cat, weights = pop_vc)

##delete rows if "suicides" == 0
library(dplyr)
rates_yearms_cyc <- rates_yearms_cyc %>%
  filter(cases != 0)
rates_yearms_cyc

##*100000
rates_yearms_cyc <- rates_yearms_cyc %>%
  mutate(across(c(rate.adj, SE.rate.adj, rate.adj.lo, rate.adj.hi, rate, SE.rate, rate.lo, rate.hi), ~ . * 100000))
rates_yearms_cyc

##table
yearms_cyc <- knitr::kable(rates_yearms_cyc, format = "html", caption = "Suicide rates by year, month and sex (per 100000)", col.names = c('year', 'month', 'sex', 'suicides', 'population', 'std_rate', 'SE std_rate', '95% LCL (std)', '95% UCL (std)', 'crude_rate', 'SE crude_rate', '95% LCL (crude)', '95% UCL (crude)'), align = 'r')
yearms_cyc


##year, month and zone
rates_yearmz_cyc <- rate(data = cyc_def1, obs = cases, pyrs = pop1, print = list(year, month, sui_zone), adjust = age_cat, weights = pop_vc)

##delete rows if "suicides" == 0
library(dplyr)
rates_yearmz_cyc <- rates_yearmz_cyc %>%
  filter(cases != 0)
rates_yearmz_cyc

##*100000
rates_yearmz_cyc <- rates_yearmz_cyc %>%
  mutate(across(c(rate.adj, SE.rate.adj, rate.adj.lo, rate.adj.hi, rate, SE.rate, rate.lo, rate.hi), ~ . * 100000))
rates_yearmz_cyc

##table
yearmz_cyc <- knitr::kable(rates_yearmz_cyc, format = "html", caption = "Suicide rates by year, month and zone (per 100000)", col.names = c('year', 'month', 'zone', 'suicides', 'population', 'std_rate', 'SE std_rate', '95% LCL (std)', '95% UCL (std)', 'crude_rate', 'SE crude_rate', '95% LCL (crude)', '95% UCL (crude)'), align = 'r')
yearmz_cyc


##year, month and age (crudes)
rates_yearma_cyc <- rate(data = cyc_def1, obs = cases, pyrs = pop1, print = list(year, month, age_cat))
rates_yearma_cyc

##*100000
rates_yearma_cyc <- rates_yearma_cyc %>%
  mutate(across(c(rate, SE.rate, rate.lo, rate.hi), ~ . * 100000))
rates_yearma_cyc

##table
yearma_cyc <- knitr::kable(rates_yearma_cyc, format = "html", caption = "Suicide rates by year, month and age groups (per 100000)", col.names = c('year', 'month', 'age', 'suicides', 'population', 'std_rate', 'SE std_rate', '95% LCL (std)', '95% UCL (std)'), align = 'r')
yearma_cyc


```

## Step 7: Confidence intervals

**Theoretical foundations**

"The number of deaths reported for any given age group and time period can be assumed to follow a **Poisson distribution**, which leads to exact confidence intervals (CIs) for age-specific mortality rates... Further, because the sum of independent Poisson random variables is Poisson-distributed, the crude mortality rate also has an exact CI. However, **no exact CI is known for age-adjusted mortality rates (i.e., directly standardized rates)**, because those are based on a weighted sum of Poisson random variables..."

**Statistics** D: number of suicides *(count: Poisson distribution).* The mean $(E)$ and variance $(V)$ satisfy:

$E(D) = \lambda P = x = V(D)$, P denotes the population denominator. The population is cancelled with the denominator of the rate, leaving the events in *t*, so the number of events $x$ is the mean and the variance.

$Crude.rate = \frac{D}{P}*100000$

$Pr(D \geq x|\lambda) = Pr(G \leq \lambda P|x) \; (eq.A1)$; the left side of the equation represents the probability of observing $x$ or more events given a $\lambda$ rate. Now, the right side represents the probability that the gamma variable $(G)$ takes values less than or equal to $\lambda P$, given $x$. This means there is an **equivalence** between the probability that there are at least $x$ events until time *t*, with a specific rate (Poisson), and the probability that the gamma random variable $(G)$ takes values less than or equal to the time *t* elapsed until $x$ events occur (cumulative time).

```{r poisson and gamma}

#cases
x <- 143

#For a range of "x" values (expected mean lambda*p)
xr <- seq(4000, 4600, by = 10)

#incidence rate
lambda <- 0.043

#persons-time
P <- 100000

#Poisson (tail complement - distribution function)
poisson_xr <- 1 - ppois(xr - 1, lambda * P)

#Gamma (distribution function)
gamma_xr <- pgamma(lambda * P, shape = xr, scale = 1)

#Results
cat("Pr(D >= x | lambda) using Poisson:", poisson_xr, "\n")
cat("Pr(G <= lambda * P | x) using Gamma:", gamma_xr, "\n")
all.equal(poisson_xr, gamma_xr) #equal

```

$G$ is gamma-distributed with shape parameter $\alpha > 0$ and scale parameter $\beta > 0$. The parameter $\alpha$ is known as the **shape parameter**, since it most influences the peakedness of the distribution, while the parameter $\beta$ is called the **scale parameter**, since most of its influence is on the spread of the distribution. The *mean* of the gamma is $\alpha\beta$; the *variance* of the gamma distribution is $\alpha\beta^2$

$$
E(G) = \alpha\beta, \quad then \quad \alpha = \frac{E(G)}{\beta} \quad (eq.1)
$$

$$
V(G) = \alpha\beta^2, \quad then \quad \alpha = \frac{V(G)}{\beta^2} \quad (eq.2)
$$

$$
eq.2 \quad in \quad eq.1: \quad \frac{V(G)}{\beta^2} = \frac{E(G)}{\beta}, \quad solving \; for \; beta \quad \beta = \frac{V(G)}{E(G)} \quad (eq.3)
$$

$$
eq.3 \quad in \quad V(G) = \alpha * \frac{V(G)^2}{E(G)^2} \quad solving \; for \; alpha \quad \alpha = \frac{E(G)^2}{V(G)} \quad (eq.4)
$$

From this equation $E(D) = \lambda P = x = V(D)$, where the expected value and variance are equal to $x$, and replacing in equations 3 and 4, we have for the gamma distribution:

$$
\alpha = x \quad and \quad \beta = 1
$$ Then:

$$
For \; the \; rate \; R = \frac{D}{P}, \; with  \; P = p, \; y = \frac{x}{p}, \; and \; v = \frac{x}{p^2}, \; the \; eq.A1 \; becomes:
$$

$$
Pr(R \geq y|\lambda) = Pr(Z \leq \lambda y|v), \; where \; Z =  \frac{G}{P} \; is \; gamma-distributed \; with \; mean \; y \; and \; variance \; v: \; \alpha = \frac{y^2}{v}; \; \beta = \frac{v}{y}
$$

From the theoretical developments of Fay-Feuer and Tiwari-Clegg-Zou:

**Lower gamma confidence limit:** $$
L(y;v) = G_{(\frac{y^2}{v},\frac{v}{y})}^{-1}(\frac{\alpha}{2}), \quad (eq.A2)
$$

**Upper gamma confidence limit (Fay-Feuer)** $$
U(y;v;W_M) = G_{(\frac{(y+W_M)^2}{v+W_M^2},\frac{v+W_M^2}{y+W_M})}^{-1}(1-\frac{\alpha}{2}), \quad (eq.A3)
$$

Note: For the upper bound, *Fay-Feuer* made a discrete approximation (by summing max. weight) to the **maximum value** that $y$ could reach; thus, the equation changes (eq.A3). Additionally, *Tiwari-Clegg-Zou* made a modification:

**Upper gamma confidence limit (Tiwari-Clegg-Zou)** $$
U(\tilde{r_i};\tilde{v_i};\bar{w_i};\alpha) = \frac{\tilde{v_i}}{2\tilde{r_i}}(\chi^2_{(2\tilde{r_i}^2/{\tilde{v_i}})})^{-1}(1-\frac{\alpha}{2}), \quad (eq.A4)
$$

This equation is expressed in the **chi-square distribution** with $2\tilde{r_i}^2/\tilde{v_i}$ degrees of freedom $(\upsilon)$. We then express it in terms of the **gamma distribution** using the following relationship: "...the chi-square distribution is just a special case of the gamma distribution...", where:

$$
\alpha = \frac{\upsilon}{2} \quad and \quad \beta = 2, \; where \; \upsilon \; is \; a \; positive \; integer
$$ Solving for alpha: $$
\upsilon = 2\tilde{r_i}^2/\tilde{v_i}, \; \alpha = \frac{\tilde{r_i}^2}{\tilde{v_i}} \; and \; \beta = 2
$$

Then: $$
U(\tilde{r_i};\tilde{v_i};\bar{w_i};\alpha) = \frac{\tilde{v_i}}{2\tilde{r_i}}(G_{(\frac{\tilde{r_i}^2}{\tilde{v_i}}, \; 2)}^{-1}(1-\frac{\alpha}{2}), \quad (eq.A5)
$$

The code below calculates confidence intervals (CIs) for standardized rates adjusted for the gamma distribution. The obtained values were compared with the intervals calculated using STATA's **"distrate"** command, which uses the gamma distribution and the methods of the authors mentioned. **The results for the rates and lower confidence intervals were the same. However, the upper confidence intervals differed slightly.**

`References:` - `Casella, G., Berger, R. L. Common Families of Distributions.  Statistical Inference. 2 ed. USA: Duxbury; 2002.` - `Talih, M., Anderson, R. N., Parker, J. D. Evaluation of four gamma-based methods for calculating confidence intervals for age-adjusted mortality rates when data are sparse. Population Health Metrics. 2022;20(1):13.` - `Fay, M. P., Feuer, E. J. Confidence intervals for directly standardized rates: a method based on the gamma distribution. Stat Med. 1997;16(7):791-801.` - `Tiwari, R. C., Clegg, L. X., Zou, Z. Efficient interval estimation for age-adjusted cancer rates. Stat Methods Med Res. 2006;15(6):547-69.` - `Department of Statistics - The Pennsylvania State University. STAT 414: Introduction to Probability Theory Pennsylvania: The Pennsylvania State University; 2024 [Available from: https://online.stat.psu.edu/stat414/lesson/15/15.8.`

```{r confidence intervals}

#by cycles!!!
alpha <- 0.05

#labels
rmod <- rates_yearm_cyc$rate.adj #rates
emod <- rates_yearm_cyc$SE.rate.adj # ~ variance 

##year and month
###gamma parameters, using 3 and 4 equations and their equivalences
shape_alpha_lo <- (rmod^2) / (emod^2)
scale_beta_lo <- (emod^2) / rmod

#of A5 equation
shape_alpha_hi <- (rmod^2) / (emod^2)
scale_beta_hi <- 2

###CIs, using A2 and A5 equations
rates_yearm_cyc$rate.adj.lo.gamma <- qgamma((alpha / 2), shape = shape_alpha_lo, scale = scale_beta_lo)
rates_yearm_cyc$rate.adj.hi.gamma <- ((emod^2)/(2*rmod)) * (qgamma((1 - (alpha / 2)), shape = shape_alpha_hi, scale = scale_beta_hi))
head(rates_yearm_cyc)

###table
yearm_cyc <- knitr::kable(rates_yearm_cyc, format = "html", caption = "Suicide rates by year and month (per 100000)", col.names = c('year', 'month', 'suicides', 'population', 'std_rate', 'SE std_rate', '95% LCL (std)', '95% UCL (std)', 'crude_rate', 'SE crude_rate', '95% LCL (crude)', '95% UCL (crude)', '95% LCL (gamma_std)', '95% UCL (gamma_std)'), align = 'r')
yearm_cyc


##year, month and age (crude rate)
###gamma parameters
shape_param <- (rates_yearma_cyc$rate^2) / (rates_yearma_cyc$SE.rate^2)
scale_param <- rates_yearma_cyc$rate / shape_param

###CIs
rates_yearma_cyc$rate.lo.gamma <- qgamma((alpha / 2), shape = shape_param, scale = scale_param)
rates_yearma_cyc$rate.hi.gamma <- qgamma((1 - (alpha / 2)), shape = shape_param, rate = (1 / scale_param))
head(rates_yearma_cyc)

###table
yearma_cyc <- knitr::kable(rates_yearma_cyc, format = "html", caption = "Suicide rates by year, month and age groups (per 100000)", col.names = c('year', 'month', 'age', 'suicides', 'population', 'rate', 'SE_rate', '95% LCL (crude)', '95% UCL (crude)', '95% LCL (gamma)', '95% UCL (gamma)'), align = 'r')
yearma_cyc


##year, month and sex
###gamma parameters
shape_param <- (rates_yearms_cyc$rate.adj^2) / (rates_yearms_cyc$SE.rate.adj^2)
scale_param <- rates_yearms_cyc$rate.adj / shape_param

###CIs
rates_yearms_cyc$rate.adj.lo.gamma <- qgamma((alpha / 2), shape = shape_param, scale = scale_param)
rates_yearms_cyc$rate.adj.hi.gamma <- qgamma((1 - (alpha / 2)), shape = shape_param, rate = (1 / scale_param))
head(rates_yearms_cyc)

###table
yearms_cyc <- knitr::kable(rates_yearms_cyc, format = "html", caption = "Suicide rates by year, month and sex (per 100000)", col.names = c('year', 'month', 'sex', 'suicides', 'population', 'std_rate', 'SE std_rate', '95% LCL (std)', '95% UCL (std)', 'crude_rate', 'SE crude_rate', '95% LCL (crude)', '95% UCL (crude)', '95% LCL (gamma_std)', '95% UCL (gamma_std)'), align = 'r')
yearms_cyc


##year, month and zone
###gamma parameters
shape_param <- (rates_yearmz_cyc$rate.adj^2) / (rates_yearmz_cyc$SE.rate.adj^2)
scale_param <- rates_yearmz_cyc$rate.adj / shape_param

###CIs
rates_yearmz_cyc$rate.adj.lo.gamma <- qgamma((alpha / 2), shape = shape_param, scale = scale_param)
rates_yearmz_cyc$rate.adj.hi.gamma <- qgamma((1 - (alpha / 2)), shape = shape_param, rate = (1 / scale_param))
head(rates_yearmz_cyc)

###table
yearmz_cyc <- knitr::kable(rates_yearmz_cyc, format = "html", caption = "Suicide rates by year, month and zone (per 100000)", col.names = c('year', 'month', 'zone', 'suicides', 'population', 'std_rate', 'SE std_rate', '95% LCL (std)', '95% UCL (std)', 'crude_rate', 'SE crude_rate', '95% LCL (crude)', '95% UCL (crude)', '95% LCL (gamma_std)', '95% UCL (gamma_std)'), align = 'r')
yearmz_cyc


```

## Step 8: Tables and final merge

To facilitate the construction of the final tables, only the columns necessary for subsequent analysis will be kept. Then, the format is adjusted **from long to wide** `(library(tidyr))`.

```{r tables}

#by decades (without CIs!!!)
##global
ymonth_dec <- rates_yearm_dec %>%
  dplyr::select(year, month, `rate.adj`)

##sex
ymsex_dec <- rates_yearms_dec %>%
  dplyr::select(year, month, sex, `rate.adj`)

ymsex_dec <- ymsex_dec %>%
  pivot_wider(names_from = sex, values_from = `rate.adj`)

##zone
ymzone_dec <- rates_yearmz_dec %>%
  dplyr::select(year, month, sui_zone, `rate.adj`)

ymzone_dec <- ymzone_dec %>%
  pivot_wider(names_from = sui_zone, values_from = `rate.adj`)

##age
ymage_dec <- rates_yearma_dec %>%
  dplyr::select(year, month, age_cat, `rate`)

ymage_dec <- ymage_dec %>%
  pivot_wider(names_from = age_cat, values_from = `rate`)

##merge (decades)
rates_std_d1 <- dplyr::left_join(ymonth_dec, ymsex_dec, by = c("year", "month"))
rates_std_d2 <- dplyr::left_join(rates_std_d1, ymzone_dec, by = c("year", "month"))
rates_std_dec <- dplyr::left_join(rates_std_d2, ymage_dec, by = c("year", "month"))


#by cycles + CIs
##global
ymonth_cyc <- rates_yearm_cyc %>%
  dplyr::select(year, month, `rate.adj`, `rate.adj.lo.gamma`, `rate.adj.hi.gamma`)

##sex
ymsex_cyc <- rates_yearms_cyc %>%
  dplyr::select(year, month, sex, `rate.adj`, `rate.adj.lo.gamma`, `rate.adj.hi.gamma`)

ymsex_cyc <- ymsex_cyc %>%
  pivot_wider(names_from = sex, values_from = c(`rate.adj`, `rate.adj.lo.gamma`, `rate.adj.hi.gamma`))

##zone
ymzone_cyc <- rates_yearmz_cyc %>%
  dplyr::select(year, month, sui_zone, `rate.adj`, `rate.adj.lo.gamma`, `rate.adj.hi.gamma`)

ymzone_cyc <- ymzone_cyc %>%
  pivot_wider(names_from = sui_zone, values_from = c(`rate.adj`, `rate.adj.lo.gamma`, `rate.adj.hi.gamma`))

##age
ymage_cyc <- rates_yearma_cyc %>%
  dplyr::select(year, month, age_cat, `rate`, `rate.lo.gamma`, `rate.hi.gamma`)

ymage_cyc <- ymage_cyc %>%
  pivot_wider(names_from = age_cat, values_from = c(`rate`, `rate.lo.gamma`, `rate.hi.gamma`))

##merge (cycles)
rates_std_c1 <- dplyr::left_join(ymonth_cyc, ymsex_cyc, by = c("year", "month"))
rates_std_c2 <- dplyr::left_join(rates_std_c1, ymzone_cyc, by = c("year", "month"))
rates_std_cyc <- dplyr::left_join(rates_std_c2, ymage_cyc, by = c("year", "month"))
rates_std_cyc

##sort the columns
rates_std_cyc <- rates_std_cyc %>%
  select(year, month, rate.adj, rate.adj.lo.gamma, rate.adj.hi.gamma, rate.adj_Women, rate.adj.lo.gamma_Women,
         rate.adj.hi.gamma_Women, rate.adj_Men, rate.adj.lo.gamma_Men, rate.adj.hi.gamma_Men, rate.adj_Rural, 
         rate.adj.lo.gamma_Rural, rate.adj.hi.gamma_Rural, rate.adj_Urban, rate.adj.lo.gamma_Urban, rate.adj.hi.gamma_Urban, 
         `rate_5-17`, `rate.lo.gamma_5-17`, `rate.hi.gamma_5-17`, `rate_18-44`, `rate.lo.gamma_18-44`, `rate.hi.gamma_18-44`, 
         `rate_45-59`, `rate.lo.gamma_45-59`, `rate.hi.gamma_45-59`, `rate_60+`, `rate.lo.gamma_60+`, `rate.hi.gamma_60+`)
rates_std_cyc


```

## Step 9: Final format and Excel document

```{r final format and Excel}

#by decades (without CIs!!!)
rates_std_dec <- rates_std_dec %>%
  rename(
    "overall" = "rate.adj",
    "women" = "Women",
    "men" = "Men",
    "rural" = "Rural",
    "urban" = "Urban"
  ) %>%
  
  mutate(
    month = case_when(
      month == "January" ~ 1,
      month == "February" ~ 2,
      month == "March" ~ 3,
      month == "April" ~ 4,
      month == "May" ~ 5,
      month == "June" ~ 6,
      month == "July" ~ 7,
      month == "August" ~ 8,
      month == "September" ~ 9,
      month == "October" ~ 10,
      month == "November" ~ 11,
      month == "December" ~ 12,
    )
  )

##Excel
library(openxlsx)
book_xlsx_dec <- createWorkbook() #create book
addWorksheet(book_xlsx_dec, "sheet1") #create sheet
writeData(book_xlsx_dec, sheet = "sheet1", x = rates_std_dec) #export table to Excel sheet
saveWorkbook(book_xlsx_dec, "rates_std_decades.xlsx", overwrite = T) #save Excel


#by cycles (with CIs)
rates_std_cyc <- rates_std_cyc %>%
  rename(
    "overall" = "rate.adj",
    "95% lci_overall" = `rate.adj.lo.gamma`,
    "95% uci_overall" = `rate.adj.hi.gamma`,
    "women" = "rate.adj_Women",
    "95% lci_women" = `rate.adj.lo.gamma_Women`,
    "95% uci_women" = `rate.adj.hi.gamma_Women`,
    "men" = "rate.adj_Men",
    "95% lci_men" = `rate.adj.lo.gamma_Men`,
    "95% uci_men" = `rate.adj.hi.gamma_Men`,
    "rural" = "rate.adj_Rural",
    "95% lci_rural" = `rate.adj.lo.gamma_Rural`,
    "95% uci_rural" = `rate.adj.hi.gamma_Rural`,
    "urban" = "rate.adj_Urban",
    "95% lci_urban" = `rate.adj.lo.gamma_Urban`,
    "95% uci_urban" = `rate.adj.hi.gamma_Urban`,
    "5-17" = "rate_5-17",
    "95% lci_5-17" = `rate.lo.gamma_5-17`,
    "95% uci_5-17" = `rate.hi.gamma_5-17`,
    "18-44" = "rate_18-44",
    "95% lci_18-44" = `rate.lo.gamma_18-44`,
    "95% uci_18-44" = `rate.hi.gamma_18-44`,
    "45-59" = "rate_45-59",
    "95% lci_45-59" = `rate.lo.gamma_45-59`,
    "95% uci_45-59" = `rate.hi.gamma_45-59`,
    "60+" = "rate_60+",
    "95% lci_60+" = `rate.lo.gamma_60+`,
    "95% uci_60+" = `rate.hi.gamma_60+`,
    
  ) %>%
  
  mutate(
    month = case_when(
      month == "January" ~ 1,
      month == "February" ~ 2,
      month == "March" ~ 3,
      month == "April" ~ 4,
      month == "May" ~ 5,
      month == "June" ~ 6,
      month == "July" ~ 7,
      month == "August" ~ 8,
      month == "September" ~ 9,
      month == "October" ~ 10,
      month == "November" ~ 11,
      month == "December" ~ 12,
    )
  )

##Excel with CIs
book_xlsx_cyc1 <- createWorkbook() 
addWorksheet(book_xlsx_cyc1, "sheet1")
writeData(book_xlsx_cyc1, sheet = "sheet1", x = rates_std_cyc)
saveWorkbook(book_xlsx_cyc1, "rates_std_cycles_ci.xlsx", overwrite = T)

##Excel without CIs
rates_std_cyc <- rates_std_cyc %>%
  select(year, month, overall, women, men, rural, urban, `5-17`, `18-44`, `45-59`, `60+`)

book_xlsx_cyc <- createWorkbook() 
addWorksheet(book_xlsx_cyc, "sheet1")
writeData(book_xlsx_cyc, sheet = "sheet1", x = rates_std_cyc)
saveWorkbook(book_xlsx_cyc, "rates_std_cycles.xlsx", overwrite = T)

#######################################################################################################################

#Print other table
##Excel
library(openxlsx)
book_xlsx_cyc1 <- createWorkbook() 
addWorksheet(book_xlsx_cyc1, "sheet1")
writeData(book_xlsx_cyc1, sheet = "sheet1", x = cyc_def1)
saveWorkbook(book_xlsx_cyc1, "cyc_def1.xlsx", overwrite = T)

```

End.
